Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-21 paper.tex
Update text for column extract figure

\subsection{Extracting, inserting and updating fields}
\begin{figure}
\includegraphics{figures/column-extract}
\caption{Time to extract the genome position and write to a text file.
Total CPU time required to extract the POS field for BCF,
sav and Zarr formats
for the data in Figure~\ref{fig-data-storage}.
For the BCF file we used \texttt{bcftools query -f"\%POS$\backslash$n"}.
For sav, we used the Savvy C++ API to extract position for each variant
and output text using the \texttt{std::cout} stream. For Zarr, we read 
the variant\_position array into a NumPy array, and then wrote to
a text file using the Pandas \texttt{write\_csv} method. 
Zarr CPU time is dominated by writing the text output; we also show
the time required to populate a NumPy array with the data in Zarr,
which is less than a second. Wall-clock time (dotted line) is dominated 
in this case by file I/O. Time to output text for Savvy is not significant
for $> 1000$ samples (not shown).
\label{fig-column-extract}}
\end{figure}
We have focused on the genotype matrix up to this point, contrasting
Zarr with existing row-wise methods.
Real-world VCFs encapsulate much more than just the genotype
matrix, and can contain large numbers of additional fields.
Fig~\ref{fig-column-extract} shows the time required to extract 
the genomic position of each variant in the simulated benchmark 
dataset, which we can use as an indicative example of a per-variant 
query. Although Savvy is many times faster than \texttt{bcftools query}
here, the row-wise storage strategy that they share means that 
the entire dataset must be read into memory and 
decompressed to extract just one field from each record. Zarr
excels at these tasks: we only read and decompress the information required.

>>>>

\subsection{Extracting, inserting and updating fields}
\begin{figure}
\includegraphics{figures/column-extract}
\caption{Time to extract the genome position and write to a text file.
Total CPU time required to extract the POS field for BCF,
sav and Zarr formats
for the data in Figure~\ref{fig-data-storage}.
For the BCF file we used \texttt{bcftools query -f"\%POS$\backslash$n"}.
For sav, we used the Savvy C++ API to extract position for each variant
and output text using the \texttt{std::cout} stream. For Zarr, we read 
the variant\_position array into a NumPy array, and then wrote to
a text file using the Pandas \texttt{write\_csv} method. 
Zarr CPU time is dominated by writing the text output; we also show
the time required to populate a NumPy array with the data in Zarr,
which is 2 seconds. Wall-clock time (dotted line) is dominated 
in this case by file I/O. Note that our benchmarking methodology of 
dropping the file system caches before each run (see Methods) is
likely not reflective of real world usage where a small and 
frequently used field such as POS would remain in cache.
Time to output text for Savvy is not significant
for $> 1000$ samples (not shown).
\label{fig-column-extract}}
\end{figure}
We have focused on the genotype matrix up to this point, contrasting
Zarr with existing row-wise methods.
Real-world VCFs encapsulate much more than just the genotype
matrix, and can contain large numbers of additional fields.
Fig~\ref{fig-column-extract} shows the time required to extract 
the genomic position of each variant in the simulated benchmark 
dataset, which we can use as an indicative example of a per-variant 
query. Although Savvy is many times faster than \texttt{bcftools query}
here, the row-wise storage strategy that they share means that 
the entire dataset must be read into memory and 
decompressed to extract just one field from each record. Zarr
excels at these tasks: we only read and decompress the information required.
